Making Higher-Order Superposition Work

### 2022-03-12 ho-techniques/ho-techniques.tex
more

>>>>

\section{Preprocessing Higher-Order Problems}
\label{sec:ho-tech:preprocessing}

The TPTP library contains thousands of higher-order problems. Despite their
diversity, they have a markedly different flavor from the TPTP first-order
problems. Notably, they extensively use the \verb|definition| role to identify
universally quantified equations (and equivalences) that define symbols.
%
Definitions $s \eq t$ (or $(s \iequiv t) \eq \itrue$) can be replaced by rewrite
rules $s \longrightarrow t$,
using the orientation given in the input problem. If there are multiple
definitions for the same symbol, only the first one is replaced by a rewrite rule.
Then, whenever a clause is picked in the given clause procedure, it will be rewritten
using the collected rules.
Alternatively, we can rewrite
the input formulas as a preprocessing step. This ensures that the input
clauses will be fully simplified when the proving process starts and no
defined symbols will occur in clauses, which usually helps the heuristics.

Since the TPTP format enforces no constraints on
definitions, rewriting might diverge. To ensure
termination, we limit the number of applied rewrite steps. In
practice, most TPTP problems are well behaved: Only one
definition is given for each symbol, and the definitions are acyclic.

Turning the defining equations into rewrite rules, unfolding the definitions, and
$\beta$-reduc\-ing the result can eliminate all of a problem's higher-order features, making
it susceptible to first-order methods. However, this can inflate the problem
beyond recognition and compromise the refutational completeness of
superposition.

\begin{exa} 
  Removing higher-order features of a problem can have adverse effects.
  Consider the TPTP problem \texttt{NUM636\^{}3}, which defines the predicate $\cst{m}$
  as $\lambda x.\, \cst{s} \, x \ineq x$ and states its conjecture as $\iforall
  x.\, \cst{m} \, x $, where $\cst{s}$ is the standard Peano-style natural number
  successor constructor. When this definition is kept as is, the
  prover can superpose from either $\cst{m}$ or its definition into the
  (clausified) induction axiom, which is also given in the problem, and quickly prove
  the conjecture, without using any advanced inductive reasoning. In contrast,
  when the definition is
  unfolded and the problem is $\beta$-reduced, both $\cst{m}$ and the
  corresponding $\lambda$-abstraction disappear, forcing the prover to guess the
  correct instantiation for the induction axiom.
\end{exa}

We describe two techniques to mitigate these issues. The first one is based on the observation that in practice,
the explosion associated with definition unfolding mostly
manifests itself on definitions of nonpredicate symbols. In some cases, it is
preferable to rely on superposition's term order and powerful simplification
engine to rewrite the proof state rather than to blindly rewrite definitions. On
the other hand, superposition's reasoning with equivalences is often inadequate
\cite{bbtv-21-full-ho-sup, gs-05-boolsup}. Thus, it makes sense to treat only
predicate definitions as rewrite rules.

The second technique aims at preserving completeness: We can try to force the term order that
parameterizes superposition to orient as many definitions as possible and rely on
demodulation to simplify the proof state. Usually, the Knuth--Bendix order (KBO)
\cite{db-1970-kbo} is used. It compares terms by first comparing their weights,
which is the sum of all the weights assigned to the symbols it contains. Given a
symbol weight assignment $\mathcal{W}$, we can update it so that it orients
acyclic definitions from left to right assuming that they are of the form $
\cst{f} \, \overline{X}_m \eq \lambda \overline{Y}_n. \, t$, where the only free
variables in $t$ are $\overline{X}_m$, no free variable is repeated or appears
applied in $t$, and $\cst{f}$ does not occur in $t$. Then we traverse the
symbols $\cst{f}$ that are defined by such equations following the dependency
relation, starting with a symbol $\cst{f}$ that does not depend on any other
defined symbol. For each $\cst{f}$, we set $\mathcal{W}(\cst{f})$ to $w + 1$,
where $w$ is the maximum weight of the right-hand sides of $\cst{f}$'s
definitions, computed using $\mathcal{W}$. By construction, for each equation
the left-hand side is heavier. Thus, the equations are orientable from left to
right.

\begin{exa} 
  Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
  in higher-order logic. More complex modal operators (such as
  implication and equivalence) are defined in terms of basic connectives (such as negation
  and disjunction). Some of the definitions present in the problems are
  $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
  (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
  variables, and basic connectives is 2, we can orient equations using
  the above described approach as follows. Starting from symbols that do not
  depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
  $\mathcal{W}(\cst{mor}) = 17$. Then, we use these values to set
  $\mathcal{W}(\cst{mimplies}) = 37$. Clearly, these weights
  enable us to orient all definitions from left to right.
\end{exa}

% Many higher-order problems, especially those generated from proof assistants,
% contain hundreds of needless axioms.
% To filter out axioms that are unlikely to be useful in a proof attempt, many
% theorem provers rely on the SInE algorithm \cite{hv-2011-sine}. SInE starts with
% the set of symbols occurring in the conjecture and tries to find axioms that
% define the properties of these symbols. Then it looks for the definitions of
% newly found symbols until it reaches a fixpoint. Axioms annotated with
% \verb|definition| ease this search because they explicitly record the
% dependency between a symbol and its characterization. To exploit
% this information, we modified SInE to optionally include the definitions of
% symbols in the conjecture, regardless of whether they are filtered out or not.
% Similarly, we implemented mode of SInE which selects only conjecture and axioms
% annotated with \verb|definition|.

\ourpara{Evaluation and Discussion}

% The \textit{base} configuration treats all axioms annotated with
% \texttt{definition} as rewrite rules applied as preprocessing.
% In addition, we tested
We designed and evaluated the following strategies for handling
\texttt{definition} axioms:

\begin{description}[labelwidth=\widthof{\rm no-RW$+$KBO~}]
  \item[\rm pre-RW~] rewrite all definitions as a preprocessing step;
  \item[\rm in-RW~] rewrite all definitions during the saturation, as an inprocessing step;
  \item[\rm $o$-RW~] rewrite only predicate definitions, during preprocessing;
  \item[\rm $o$-RW$+$KBO~] like $o$-RW but with adjusted KBO weights for the remaining
    definitions;
  \item[\rm no-RW~] no special treatment of definitions;
  \item[\rm no-RW$+$KBO~] like $no$-RW but adjusting KBO weights for all definitions.
\end{description}

The results are given in Fig.~\ref{fig:rewrite}. In all the figures, each
cell gives the number of proved problems, and cells marked with $\star$
correspond to the base configuration. The highest number in a category is typeset in
\relax{bold}. SH benchmarks are not
included because they do not contain the \texttt{definition} role.

\newcommand{\unknownres}{\ensuremath{{\varnothing}}}
\newcommand{\colalign}{\phantom{0}}

\begin{figure}[t]
  \centering
  \def\arraystretch{1.1}%
  \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
                & pre-RW                 & in-RW        & $o$-RW               & $o$-RW$+$KBO    & no-RW        & no-RW$ + $KBO  \\ \midrule
           TPTP & {\bf 1635}$^\star$     & 1619         & 1620                 & 1621            & 1298         & 1296
           \\ \bottomrule
         \end{tabular}}
       \captionof{figure}{Impact of the definition rewriting method}
       \label{fig:rewrite}
     \end{figure}

The four configurations in which definitions are treated as rewrite rules
perform much better than the other two. In contrast, adjusting KBO weights gives
no substantial improvement: Looking at raw data, we found only
\NumberOK{2}~problems proved by $o$-RW$+$KBO but not by $o$-RW in which the
feature was used in the proof. For no-RW and no-RW$+$KBO, the
\NumberOK{2}-problem difference may be just noise. Even though it proves fewer
problems, the configuration $o$-RW has some advantages over pre-RW: It proves
\NumberOK{16} problems that pre-RW does not, \NumberOK{3} of which have a TPTP
difficulty rating (the ratio of eligible %TPTP
provers that cannot prove the problem) of~1.

% \looseness=-1
Rewriting after clausification avoids getting stuck rewriting parts of the
proof state that might not contribute to the proof. In practice, we noticed that
rewriting can be so expensive that the prover can spend all
allotted CPU time in the preprocessing phase. The evaluation results confirm this
observation: There are \NumberOK{64} problems proved by in-RW but not by
pre-RW. Moreover, there are \NumberOK{41} problems that can be proved only
by in-RW but not by any other above described configuration. % from Fig.~\ref{fig:rewrite}.

\section{Reasoning about Formulas}
\label{sec:ho-tech:formulas}

Higher-order logic identifies formulas with terms of Boolean type. To prove a problem, we often
need to instantiate a variable with the right predicate.
Finding this predicate can be easier if the problem is not clausified.
Consider the conjecture $\iexists f. \, f \, \cst{p} \, \cst{q} \iequiv \cst{p}
\iand \cst{q}$. Expressed in this form, the formula is easy to prove by taking
$f := \lambda x \, y. \> x  \iand y$. By contrast, guessing the right
instantiation for the negated, clausified form $ \neglit{F \, \cst{p} \,
\cst{q}} \llor \neglit{\cst{p}} \llor \neglit{\cst{q}},
\poslit{F \, \cst{p} \, \cst{q}} \llor \poslit{\cst{p}}$, $\poslit{F \, \cst{p}
\, \cst{q}} \llor \poslit{\cst{q}}$ is more challenging.
One of the strengths of higher-order tableau provers is that they do not clausify the input
problem. This might partly explain Satallax's dominance in the THF division of CASC
competitions until the 2020 edition of CASC.

The \osup{} calculus supports \emph{delayed clausification rules} that insert
problems into the proof state in their original, nonclausified form, and
clausify them gradually. Delayed clausification allows the prover to analyze the
syntactic structure of formulas during saturation, whereas the more traditional
approach of \emph{immediate clausification} applies a standard clausification
algorithm \cite{nw-01-small-cnf} both as a preprocessing step and whenever
predicate variables are instantiated.

An earlier evaluation of the \osup{} calculus \cite{bbtv-21-full-ho-sup} showed
that the \emph{outer} variant of delayed clausification substantially increases
this calculus's performance. The outer variant clausifies top-level logical
symbols, proceeding from the outside inwards; this method corresponds to the one
described in the previous chapter. For example, a clause $C \llor
\neglit{(\cst{p} \iand \cst{q})}$ is transformed into $C \llor \neglit{\cst p}
\llor \neglit{\cst q}$. The calculus also supports \emph{inner} delayed
clausification, which uses only the core calculus rules to clausify problems.
Even though this is the laziest approach to clausification, the earlier
evaluation by Bentkamp et al. \cite{bbtv-21-full-ho-sup} showed that this
approach is inefficient. Thus, we focus only on the outer rules.

\looseness=-1
Delayed clausification rules can be used as inference rules (which add conclusions
to the passive set) or as simplification rules (which delete premises and add
conclusions to the passive set).
%
Inferences are more flexible, as all
intermediate clausification states will be stored in the proof state, at the
cost of producing many clauses. Simplifications produce fewer clauses,
but risk destroying informative syntactic structure.
Since clausifying equivalences can destroy a lot of syntactic structure
\cite{gs-05-boolsup}, we never apply simplifying rules
on them.

Delayed clausification can interfere with clause splitting techniques.
Zipperposition supports a lightweight variant of AVATAR \cite{av-2014-avatar},
an architecture that partitions the search space by
splitting clauses into variable-disjoint subclauses. This lightweight AVATAR is described
by Ebner et al.\ \cite[Sect.~7]{2021-ebt-unifying-splitting}. Combining it
with delayed clausification makes it possible to split a %higher-order
clause $(\varphi_1 \ior \cdots \ior \varphi_n) \eq \itrue$, where
the $\varphi_i$'s are arbitrarily complex formulas that share no free
variables with each other, into clauses $\varphi_i \eq \itrue$.
%
To finish the proof, it suffices to derive the empty clause under each assumption
$\varphi_i \eq \itrue$. Since the split is performed at the formula level, this
technique resembles tableaux, but it exploits the strengths of superposition,
such as its powerful redundancy criterion and simplification machinery, to
close the branches.

### 2022-03-12 scratch/paper.tex
more

\spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}
\spnewtheorem{examplex}[theorem]{Example}{\bfseries}{\rmfamily}

@@

\begin{examplex} 
  Removing higher-order features of a problem can have adverse effects.
  Consider the TPTP problem \texttt{NUM636\^{}3}, which defines the predicate $\cst{m}$
  as $\lambda x.\, \cst{s} \, x \ineq x$ and states its conjecture as $\iforall
  x.\, \cst{m} \, x $, where $\cst{s}$ is the standard Peano-style natural number
  successor constructor. When this definition is kept as is, the
  prover can superpose from either $\cst{m}$ or its definition into the
  (clausified) induction axiom, which is also given in the problem, and quickly prove
  the conjecture, without using any advanced inductive reasoning. In contrast,
  when the definition is
  unfolded and the problem is $\beta$-reduced, both $\cst{m}$ and the
  corresponding $\lambda$-abstraction disappear, forcing the prover to guess the
  correct instantiation for the induction axiom.
\end{examplex}

@@

\begin{examplex} 
  Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
  in higher-order logic. More complex modal operators (such as
  implication and equivalence) are defined in terms of basic connectives (such as negation
  and disjunction). Some of the definitions present in the problems are
  $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
  (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
  variables, and basic connectives is 2, we can orient equations using
  the above described approach as follows. Starting from symbols that do not
  depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
  $\mathcal{W}(\cst{mor}) = 17$. Then, we use these values to set
  $\mathcal{W}(\cst{mimplies}) = 37$. Clearly, these weights
  enable us to orient all definitions from left to right.
\end{examplex}

@@

\begin{examplex}
\label{ex:dat056-2}
The clausified conjecture of the problem \texttt{DAT056\^{}2}
\cite{ns-13-leo2sh} from the TPTP library is $\cst{ap} \, \cst{xs}
\, (\cst{ap} \, \cst{ys} \, \cst{zs}) \not\eq \cst{ap} \, (\cst{ap} \, \cst{xs} \,
\cst{ys}) \, \cst{zs}$, where $\cst{ap}$ is the \confrep{}{list }append operator defined
recursively on its first argument and $\cst{xs}$, $\cst{ys}$, and $\cst{zs}$ are
of list type. Abstracting $\cst{xs}$ from the disequation yields $t = \lambda \mathit{xs}.\, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
\allowbreak\ieq\allowbreak \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$, which is added
to $\mathit{Inst}$.
Included in the problem is the induction axiom
for the list datatype: $\iforall p. \, p \, \cst{nil} \iand (\iforall x \,
\mathit{xs}. \, p \, \mathit{xs} \iimplies\allowbreak p \, (\cst{cons} \, x \,
\mathit{xs})) \iimplies\allowbreak \iforall \mathit{xs}. \, p \, \mathit{xs}$, where
$\cst{nil}$ and $\cst{cons}$ have the usual meanings.
Instantiating $p$ with $t$ and
using the $\cst{ap}$ definition, we can prove
$\iforall \mathit{xs}. \, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
\ieq \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$,
from which we easily derive a contradiction.
\end{examplex}

@@

\begin{examplex} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
(\iforall \, (\lambda x. \, \cst{q} \, x)) \, \cst{b}$ are the term itself, $F \, \cst{a}$, $\cst{p} \, (\iforall \, (\lambda x. \, \cst{q} \, x))
\, \cst{b}$, $\iforall \, (\lambda x. \, \cst{q} \, x)$, and $\cst{b}$.
\end{examplex} 
%
Green positions are lifted to clauses as follows: If $p$ is the green position
of a subterm in $s$, and $s$ occurs in a literal $l \in \{s \eq t{,}\; s \not\eq
t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
\osup{} mandates additional restrictions on the Boolean selection function:
$\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
$s \eq t$, neither $s$ not $t$ cannot be selected; if a term $s$ contains a
variable $X$ as a green subterm, and $X\> \tuplen{u}$, with $n \ge 1$, is a
maximal term of the clause, $s$ cannot be selected.

@@

\begin{examplex}
  Selecting the right Boolean subterm can help avoid elaborating
  high\-er-order inferences. Consider the unsatisfiable clause set consisting of $\cst{p} \, (\lambda y.\, X
  \, (\lambda x.\, x) \, \cst{a}) \iimplies\allowbreak \inot(\cst{p} \, (\lambda y. \allowbreak\, X
  \, y \, \cst{a}))$, $\cst{p} \, (\lambda y.\, \cst{a})$, and
  $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$.
  Note that $\cst{p} \, (\lambda y.\, X
  \, (\lambda x.\, x) \, \cst{a})$ and $\cst{p} \,
  (\lambda y.\, \cst{a})$ have infinitely many unifiers of the form $\{ X \mapsto \lambda fx.
  \, f^i\, (x) \}, i \geq 0$, whereas $\cst{p} \, (\lambda y. \, X
  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have only one unifier. 
  If \texttt{Forward} context selection is enabled,
  $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$ is made the target of superposition inference, 
  forcing computation of at least 100 unifiers (under the assumption that
  unifiers are returned in order of increasing $i$) before we get to refute
  $\inot(\cst{p} \, (\lambda y. y^{100} \, \cst{b}))$.
  In contrast, \texttt{Backward} context selection allows us to
  superpose from $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$ into $\cst{p} \, (\lambda y. \, X
  \, y \, \cst{a})$, avoiding this explosion. 
\end{examplex}

@@

\begin{examplex} 
  \begin{sloppypar}
  Consider the unsatisfiable two-clause problem $\{ X \, (\cst{f} \,
  \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}),\allowbreak
  \neg \cst{p} \, (\cst{f}^{100} \, \cst{a})  \}$ and a selection function which
  selects negative literals.
  Let $P \mid A \mid Q$ denote
  the state of the given clause loop (i.e., the contents of the passive and active set
  and of the stream queue), and let $\infstream{ a_1, a_2, \ldots }$
  denote an infinite stream of elements.
  \end{sloppypar}

@@

  In the next iterations, some clause of the form $\cst{p} \, (\cst{f}^{i} \,
  \cst{a})$, where $i < 100$, is chosen, but no inferences with it can be
  performed. Then, the stream created in the second iteration is probed, and its
  results fill the set $P$. Ultimately, the clause $\cst{p} \, (\cst{f}^{100} \,
  \cst{a})$ is chosen, at which point $\bot$ is quickly derived.
\end{examplex}

>>>>

\spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}
\spnewtheorem{exa}[theorem]{Example}{\bfseries}{\rmfamily}

@@

\begin{exa} 
  Removing higher-order features of a problem can have adverse effects.
  Consider the TPTP problem \texttt{NUM636\^{}3}, which defines the predicate $\cst{m}$
  as $\lambda x.\, \cst{s} \, x \ineq x$ and states its conjecture as $\iforall
  x.\, \cst{m} \, x $, where $\cst{s}$ is the standard Peano-style natural number
  successor constructor. When this definition is kept as is, the
  prover can superpose from either $\cst{m}$ or its definition into the
  (clausified) induction axiom, which is also given in the problem, and quickly prove
  the conjecture, without using any advanced inductive reasoning. In contrast,
  when the definition is
  unfolded and the problem is $\beta$-reduced, both $\cst{m}$ and the
  corresponding $\lambda$-abstraction disappear, forcing the prover to guess the
  correct instantiation for the induction axiom.
\end{exa}

@@

\begin{exa} 
  Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
  in higher-order logic. More complex modal operators (such as
  implication and equivalence) are defined in terms of basic connectives (such as negation
  and disjunction). Some of the definitions present in the problems are
  $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
  (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
  variables, and basic connectives is 2, we can orient equations using
  the above described approach as follows. Starting from symbols that do not
  depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
  $\mathcal{W}(\cst{mor}) = 17$. Then, we use these values to set
  $\mathcal{W}(\cst{mimplies}) = 37$. Clearly, these weights
  enable us to orient all definitions from left to right.
\end{exa}

@@

\begin{exa}
\label{ex:dat056-2}
The clausified conjecture of the problem \texttt{DAT056\^{}2}
\cite{ns-13-leo2sh} from the TPTP library is $\cst{ap} \, \cst{xs}
\, (\cst{ap} \, \cst{ys} \, \cst{zs}) \not\eq \cst{ap} \, (\cst{ap} \, \cst{xs} \,
\cst{ys}) \, \cst{zs}$, where $\cst{ap}$ is the \confrep{}{list }append operator defined
recursively on its first argument and $\cst{xs}$, $\cst{ys}$, and $\cst{zs}$ are
of list type. Abstracting $\cst{xs}$ from the disequation yields $t = \lambda \mathit{xs}.\, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
\allowbreak\ieq\allowbreak \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$, which is added
to $\mathit{Inst}$.
Included in the problem is the induction axiom
for the list datatype: $\iforall p. \, p \, \cst{nil} \iand (\iforall x \,
\mathit{xs}. \, p \, \mathit{xs} \iimplies\allowbreak p \, (\cst{cons} \, x \,
\mathit{xs})) \iimplies\allowbreak \iforall \mathit{xs}. \, p \, \mathit{xs}$, where
$\cst{nil}$ and $\cst{cons}$ have the usual meanings.
Instantiating $p$ with $t$ and
using the $\cst{ap}$ definition, we can prove
$\iforall \mathit{xs}. \, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
\ieq \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$,
from which we easily derive a contradiction.
\end{exa}

@@

\begin{exa} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
(\iforall \, (\lambda x. \, \cst{q} \, x)) \, \cst{b}$ are the term itself, $F \, \cst{a}$, $\cst{p} \, (\iforall \, (\lambda x. \, \cst{q} \, x))
\, \cst{b}$, $\iforall \, (\lambda x. \, \cst{q} \, x)$, and $\cst{b}$.
\end{exa} 
%
Green positions are lifted to clauses as follows: If $p$ is the green position
of a subterm in $s$, and $s$ occurs in a literal $l \in \{s \eq t{,}\; s \not\eq
t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
\osup{} mandates additional restrictions on the Boolean selection function:
$\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
$s \eq t$, neither $s$ not $t$ cannot be selected; if a term $s$ contains a
variable $X$ as a green subterm, and $X\> \tuplen{u}$, with $n \ge 1$, is a
maximal term of the clause, $s$ cannot be selected.

@@

\begin{exa}
  Selecting the right Boolean subterm can help avoid elaborating
  high\-er-order inferences. Consider the unsatisfiable clause set consisting of $\cst{p} \, (\lambda y.\, X
  \, (\lambda x.\, x) \, \cst{a}) \iimplies\allowbreak \inot(\cst{p} \, (\lambda y. \allowbreak\, X
  \, y \, \cst{a}))$, $\cst{p} \, (\lambda y.\, \cst{a})$, and
  $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$.
  Note that $\cst{p} \, (\lambda y.\, X
  \, (\lambda x.\, x) \, \cst{a})$ and $\cst{p} \,
  (\lambda y.\, \cst{a})$ have infinitely many unifiers of the form $\{ X \mapsto \lambda fx.
  \, f^i\, (x) \}, i \geq 0$, whereas $\cst{p} \, (\lambda y. \, X
  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have only one unifier. 
  If \texttt{Forward} context selection is enabled,
  $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$ is made the target of superposition inference, 
  forcing computation of at least 100 unifiers (under the assumption that
  unifiers are returned in order of increasing $i$) before we get to refute
  $\inot(\cst{p} \, (\lambda y. y^{100} \, \cst{b}))$.
  In contrast, \texttt{Backward} context selection allows us to
  superpose from $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$ into $\cst{p} \, (\lambda y. \, X
  \, y \, \cst{a})$, avoiding this explosion. 
\end{exa}

@@

\begin{exa} 
  \begin{sloppypar}
  Consider the unsatisfiable two-clause problem $\{ X \, (\cst{f} \,
  \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}),\allowbreak
  \neg \cst{p} \, (\cst{f}^{100} \, \cst{a})  \}$ and a selection function which
  selects negative literals.
  Let $P \mid A \mid Q$ denote
  the state of the given clause loop (i.e., the contents of the passive and active set
  and of the stream queue), and let $\infstream{ a_1, a_2, \ldots }$
  denote an infinite stream of elements.
  \end{sloppypar}

@@

  In the next iterations, some clause of the form $\cst{p} \, (\cst{f}^{i} \,
  \cst{a})$, where $i < 100$, is chosen, but no inferences with it can be
  performed. Then, the stream created in the second iteration is probed, and its
  results fill the set $P$. Ultimately, the clause $\cst{p} \, (\cst{f}^{100} \,
  \cst{a})$ is chosen, at which point $\bot$ is quickly derived.
\end{exa}
